---
title: "Increasing Area Effect"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Increasing Area Effect}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::knitr}
editor_options: 
  markdown: 
    wrap: 100
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# How chooseGCM selection changes with increasing area?

This Rmarkdown is part of the following article:

Esser, L.F., Bailly, D., Lima, M.R., Ré, R. 2025. chooseGCM: A Toolkit to Select General Circulation Models in R. Global Change Biology , 31(1), e70008. Available at: https://doi.org/10.1111/gcb.70008.

## Introduction

chooseGCM is a solution for GCMs selection in Climate Change research. We built this Rmarkdown as a way to test the properties of the methods underlying chooseGCM. Results from each function will be presented side by side with changing variables, allowing better comparison. Here, we have already downloaded GCMs data.

This Rmarkdown will apply different algorithms implemented in chooseGCM in an increasing area. The aim is to understand from which point the GCM selection starts to change. For that, we will use *Araucaria angustifolia* records and will create a buffer from increassing size.

```{r libs, message=F, warning=F}
# Install and open CRAN libraries
for(x in c("devtools", "geodata", "terra", "tictoc") ){
  if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
  }
}

# Install and open chooseGCM
if (!require("chooseGCM", character.only = TRUE)) {
    devtools::install_github("luizesser/chooseGCM")
    library("chooseGCM", character.only = TRUE)
}

# Install and open caretSDM
if (!require("caretSDM", character.only = TRUE)) {
    devtools::install_github("luizesser/caretSDM")
    library("caretSDM", character.only = TRUE)
}
```

## Preparing data

### Download GCMs

To obtain GCMs data, we recommend using `geodata` package, which links directly to the WorldClim 2.1 database. We have previously downloaded GCMs data, but an example code is supplied below.

```{r download_gcms, eval=FALSE}
gcms <- geodata:::.cmods
for (r in c(10)) {
  for ( g in gcms ) {
    geodata::cmip6_world(model = g,
                         ssp = "585",
                         time = "2081-2100",
                         var = "bioc",
                         res = r,
                         path = ""
    )
  }
}
```

### Import GCMs

The function `import_gcms` helps the user to import data in the correct way, which is a list of GCMs. Each element of the list is a `SpatRaster` from `terra` package, containing all variables. The list should be named with GCMs names to better plot results.

```{r import_gcms10}
tictoc::tic()
s10 <- chooseGCM::import_gcms(path = "~/storage/WC_data/WC_data_all_gcms_10/", 
                              gcm_names = c("ACCESS-CM2","ACCESS-ESM1-5",
                                            "CanESM5-CanOE","CMCC-ESM2",
                                            "CNRM-CM6-1-HR","CNRM-CM6-1",
                                            "CNRM-ESM2-1","EC-Earth3-Veg",
                                            "EC-Earth3-Veg-LR","FIO-ESM-2-0",
                                            "GISS-E2-1-G","GISS-E2-1-H",
                                            "HadGEM3-GC31-LL","INM-CM5-0",
                                            "INM-CM4-8","IPSL-CM6A-LR",
                                            "MIROC-ES2L","MIROC6",
                                            "MPI-ESM1-2-LR","MPI-ESM1-2-HR",
                                            "MRI-ESM2-0","UKESM1-0-LL"))
tictoc::toc()
```

## Species Data

For this demonstration, we will use Araucaria angustifolia, the brazilian pine, as a study case. This flagship species is widely modeled and know, and is a key species for the Atlantic Rainforest biodiversity hotspot. Species records were obtained from GBIF and cleaned by excluding records from outside the native region and NAs. This set of records was then used to build 10 buffers from growing size, each buffer 1 degree-wide bigger.

```{r buffers}
occ_data <- geodata::sp_occurrence("Araucaria", "angustifolia")
occ_data <- occ_data |>
  dplyr::filter(lon >= -64) |> 
  dplyr::filter(lon <= -33) |> 
  dplyr::filter(lat <= -19.5) |> 
  dplyr::filter(lat >= -33)

buffer50 <- na.omit(occ_data[,c("lon", "lat")]) |>
  sf::st_as_sf(coords= c(1,2), crs = 4326) |> 
  sf::st_transform(crs=sf::st_crs(6933)) |>
  sf::st_buffer(dist = 50000) |> 
  sf::st_union()
buffer100 <- na.omit(occ_data[,c("lon", "lat")]) |>
  sf::st_as_sf(coords= c(1,2), crs = 4326) |> 
  sf::st_transform(crs=sf::st_crs(6933)) |>
  sf::st_buffer(dist = 100000) |> 
  sf::st_union()
buffer150 <- na.omit(occ_data[,c("lon", "lat")]) |>
  sf::st_as_sf(coords= c(1,2), crs = 4326) |> 
  sf::st_transform(crs=sf::st_crs(6933)) |>
  sf::st_buffer(dist = 150000) |> 
  sf::st_union()
buffer200 <- na.omit(occ_data[,c("lon", "lat")]) |>
  sf::st_as_sf(coords= c(1,2), crs = 4326) |> 
  sf::st_transform(crs=sf::st_crs(6933)) |>
  sf::st_buffer(dist = 200000) |> 
  sf::st_union()
buffer250 <- na.omit(occ_data[,c("lon", "lat")]) |>
  sf::st_as_sf(coords= c(1,2), crs = 4326) |> 
  sf::st_transform(crs=sf::st_crs(6933)) |>
  sf::st_buffer(dist = 250000) |> 
  sf::st_union()
buffer300 <- na.omit(occ_data[,c("lon", "lat")]) |>
  sf::st_as_sf(coords= c(1,2), crs = 4326) |> 
  sf::st_transform(crs=sf::st_crs(6933)) |>
  sf::st_buffer(dist = 300000) |> 
  sf::st_union()
buffer350 <- na.omit(occ_data[,c("lon", "lat")]) |>
  sf::st_as_sf(coords= c(1,2), crs = 4326) |> 
  sf::st_transform(crs=sf::st_crs(6933)) |>
  sf::st_buffer(dist = 350000) |> 
  sf::st_union()
buffer400 <- na.omit(occ_data[,c("lon", "lat")]) |>
  sf::st_as_sf(coords= c(1,2), crs = 4326) |> 
  sf::st_transform(crs=sf::st_crs(6933)) |>
  sf::st_buffer(dist = 400000) |> 
  sf::st_union()
buffer450 <- na.omit(occ_data[,c("lon", "lat")]) |>
  sf::st_as_sf(coords= c(1,2), crs = 4326) |> 
  sf::st_transform(crs=sf::st_crs(6933)) |>
  sf::st_buffer(dist = 450000) |> 
  sf::st_union()
buffer500 <- na.omit(occ_data[,c("lon", "lat")]) |>
  sf::st_as_sf(coords= c(1,2), crs = 4326) |> 
  sf::st_transform(crs=sf::st_crs(6933)) |>
  sf::st_buffer(dist = 500000) |> 
  sf::st_union()
plot(buffer50, axes=T)
```

## Applying algorithms

First we will set a seed to standardize randomic outputs.

```{r seed}
set.seed(1)
```

Now we will perform the Monte Carlo analysis to each type of algorithm and to different sets of variables.

### Hierarchical Clustering selection

```{r hclust_gcms50}
tictoc::tic("Hierarchical Clustering 50")
mc_hclust_50 <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer50, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "hclust")
mc_hclust_50
hclust_50 <- mc_hclust_50$suggested_gcms$k3
tictoc::toc()
```

```{r hclust_gcms100}
tictoc::tic("Hierarchical Clustering 100")
mc_hclust_100 <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer100, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "hclust")
mc_hclust_100
hclust_100 <- mc_hclust_100$suggested_gcms$k3
tictoc::toc()
```

```{r hclust_gcms150}
tictoc::tic("Hierarchical Clustering 150")
mc_hclust_150 <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer150, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "hclust")
mc_hclust_150
hclust_150 <- mc_hclust_150$suggested_gcms$k3
tictoc::toc()
```

```{r hclust_gcms200}
tictoc::tic("Hierarchical Clustering 200")
mc_hclust_200 <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer200, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "hclust")
mc_hclust_200
hclust_200 <- mc_hclust_200$suggested_gcms$k3
tictoc::toc()
```

```{r hclust_gcms250}
tictoc::tic("Hierarchical Clustering 250")
mc_hclust_250 <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer250, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "hclust")
mc_hclust_250
hclust_250 <- mc_hclust_250$suggested_gcms$k3
tictoc::toc()
```

```{r hclust_gcms300}
tictoc::tic("Hierarchical Clustering 300")
mc_hclust_300 <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer300, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "hclust")
mc_hclust_300
hclust_300 <- mc_hclust_300$suggested_gcms$k3
tictoc::toc()
```

```{r hclust_gcms350}
tictoc::tic("Hierarchical Clustering 350")
mc_hclust_350 <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer350, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "hclust")
mc_hclust_350
hclust_350 <- mc_hclust_350$suggested_gcms$k3
tictoc::toc()
```

```{r hclust_gcms400}
tictoc::tic("Hierarchical Clustering 400")
mc_hclust_400 <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer400, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "hclust")
mc_hclust_400
hclust_400 <- mc_hclust_400$suggested_gcms$k3
tictoc::toc()
```

```{r hclust_gcms450}
tictoc::tic("Hierarchical Clustering 450")
mc_hclust_450 <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer450, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "hclust")
mc_hclust_450
hclust_450 <- mc_hclust_450$suggested_gcms$k3
tictoc::toc()
```

```{r hclust_gcms500}
tictoc::tic("Hierarchical Clustering 500")
mc_hclust_500 <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer500, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "hclust")
mc_hclust_500
hclust_500 <- mc_hclust_500$suggested_gcms$k3
tictoc::toc()
```


### K-means Clustering selection

```{r kmeans_gcms50}
tictoc::tic("K-means Clustering 50")
mc_kmeans_50 <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer50, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "kmeans")
mc_kmeans_50
kmeans_50 <- mc_kmeans_50$suggested_gcms$k3
tictoc::toc()
```

```{r kmeans_gcms100}
tictoc::tic("K-means Clustering 100")
mc_kmeans_100 <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer100, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "kmeans")
mc_kmeans_100
kmeans_100 <- mc_kmeans_100$suggested_gcms$k3
tictoc::toc()
```

```{r kmeans_gcms150}
tictoc::tic("K-means Clustering 150")
mc_kmeans_150 <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer150, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "kmeans")
mc_kmeans_150
kmeans_150 <- mc_kmeans_150$suggested_gcms$k3
tictoc::toc()
```

```{r kmeans_gcms200}
tictoc::tic("K-means Clustering 200")
mc_kmeans_200 <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer200, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "kmeans")
mc_kmeans_200
kmeans_200 <- mc_kmeans_200$suggested_gcms$k3
tictoc::toc()
```

```{r kmeans_gcms250}
tictoc::tic("K-means Clustering 250")
mc_kmeans_250 <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer250, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "kmeans")
mc_kmeans_250
kmeans_250 <- mc_kmeans_250$suggested_gcms$k3
tictoc::toc()
```

```{r kmeans_gcms300}
tictoc::tic("K-means Clustering 300")
mc_kmeans_300 <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer300, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "kmeans")
mc_kmeans_300
kmeans_300 <- mc_kmeans_300$suggested_gcms$k3
tictoc::toc()
```

```{r kmeans_gcms350}
tictoc::tic("K-means Clustering 350")
mc_kmeans_350 <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer350, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "kmeans")
mc_kmeans_350
kmeans_350 <- mc_kmeans_350$suggested_gcms$k3
tictoc::toc()
```

```{r kmeans_gcms400}
tictoc::tic("K-means Clustering 400")
mc_kmeans_400 <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer400, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "kmeans")
mc_kmeans_400
kmeans_400 <- mc_kmeans_400$suggested_gcms$k3
tictoc::toc()
```

```{r kmeans_gcms450}
tictoc::tic("K-means Clustering 450")
mc_kmeans_450 <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer450, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "kmeans")
mc_kmeans_450
kmeans_450 <- mc_kmeans_450$suggested_gcms$k3
tictoc::toc()
```

```{r kmeans_gcms500}
tictoc::tic("K-means Clustering 500")
mc_kmeans_500 <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer500, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "kmeans")
mc_kmeans_500
kmeans_500 <- mc_kmeans_500$suggested_gcms$k3
tictoc::toc()
```


### Closestdist selection

```{r closestdist_gcms50}
tictoc::tic("Closestdist 50")
mc_closestdist_50 <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer50, 
                             method = "euclidean")
mc_closestdist_50
closestdist_50 <- mc_closestdist_50$suggested_gcms
tictoc::toc()
```

```{r closestdist_gcms100}
tictoc::tic("Closestdist 100")
mc_closestdist_100 <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer100, 
                             method = "euclidean")
mc_closestdist_100
closestdist_100 <- mc_closestdist_100$suggested_gcms
tictoc::toc()
```

```{r closestdist_gcms150}
tictoc::tic("Closestdist 150")
mc_closestdist_150 <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer150, 
                             method = "euclidean")
mc_closestdist_150
closestdist_150 <- mc_closestdist_150$suggested_gcms
tictoc::toc()
```

```{r closestdist_gcms200}
tictoc::tic("Closestdist 200")
mc_closestdist_200 <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer200, 
                             method = "euclidean")
mc_closestdist_200
closestdist_200 <- mc_closestdist_200$suggested_gcms
tictoc::toc()
```

```{r closestdist_gcms250}
tictoc::tic("Closestdist 250")
mc_closestdist_250 <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer250, 
                             method = "euclidean")
mc_closestdist_250
closestdist_250 <- mc_closestdist_250$suggested_gcms
tictoc::toc()
```

```{r closestdist_gcms300}
tictoc::tic("Closestdist 300")
mc_closestdist_300 <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer300, 
                             method = "euclidean")
mc_closestdist_300
closestdist_300 <- mc_closestdist_300$suggested_gcms
tictoc::toc()
```

```{r closestdist_gcms350}
tictoc::tic("Closestdist 350")
mc_closestdist_350 <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer350, 
                             method = "euclidean")
mc_closestdist_350
closestdist_350 <- mc_closestdist_350$suggested_gcms
tictoc::toc()
```

```{r closestdist_gcms400}
tictoc::tic("Closestdist 400")
mc_closestdist_400 <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer400, 
                             method = "euclidean")
mc_closestdist_400
closestdist_400 <- mc_closestdist_400$suggested_gcms
tictoc::toc()
```

```{r closestdist_gcms450}
tictoc::tic("Closestdist 450")
mc_closestdist_450 <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer450, 
                             method = "euclidean")
mc_closestdist_450
closestdist_450 <- mc_closestdist_450$suggested_gcms
tictoc::toc()
```

```{r closestdist_gcms500}
tictoc::tic("Closestdist 500")
mc_closestdist_500 <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer500, 
                             method = "euclidean")
mc_closestdist_500
closestdist_500 <- mc_closestdist_500$suggested_gcms
tictoc::toc()
```


### Closestdist selection

```{r closestdist3_gcms50}
tictoc::tic("Closestdist3 50")
mc_closestdist3_50 <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer50, 
                             method = "euclidean",
                             k = 3)
mc_closestdist3_50
closestdist3_50 <- mc_closestdist3_50$suggested_gcms
tictoc::toc()
```

```{r closestdist3_gcms100}
tictoc::tic("Closestdist3 100")
mc_closestdist3_100 <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer100, 
                             method = "euclidean",
                             k = 3)
mc_closestdist3_100
closestdist3_100 <- mc_closestdist3_100$suggested_gcms
tictoc::toc()
```

```{r closestdist3_gcms150}
tictoc::tic("Closestdist3 150")
mc_closestdist3_150 <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer150, 
                             method = "euclidean",
                             k = 3)
mc_closestdist3_150
closestdist3_150 <- mc_closestdist3_150$suggested_gcms
tictoc::toc()
```

```{r closestdist3_gcms200}
tictoc::tic("Closestdist3 200")
mc_closestdist3_200 <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer200, 
                             method = "euclidean",
                             k = 3)
mc_closestdist3_200
closestdist3_200 <- mc_closestdist3_200$suggested_gcms
tictoc::toc()
```

```{r closestdist3_gcms250}
tictoc::tic("Closestdist3 250")
mc_closestdist3_250 <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer250, 
                             method = "euclidean",
                             k = 3)
mc_closestdist3_250
closestdist3_250 <- mc_closestdist3_250$suggested_gcms
tictoc::toc()
```

```{r closestdist3_gcms300}
tictoc::tic("Closestdist3 300")
mc_closestdist3_300 <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer300, 
                             method = "euclidean",
                             k = 3)
mc_closestdist3_300
closestdist3_300 <- mc_closestdist3_300$suggested_gcms
tictoc::toc()
```

```{r closestdist3_gcms350}
tictoc::tic("Closestdist3 350")
mc_closestdist3_350 <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer350, 
                             method = "euclidean",
                             k = 3)
mc_closestdist3_350
closestdist3_350 <- mc_closestdist3_350$suggested_gcms
tictoc::toc()
```

```{r closestdist3_gcms400}
tictoc::tic("Closestdist3 400")
mc_closestdist3_400 <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer400, 
                             method = "euclidean",
                             k = 3)
mc_closestdist3_400
closestdist3_400 <- mc_closestdist3_400$suggested_gcms
tictoc::toc()
```

```{r closestdist3_gcms450}
tictoc::tic("Closestdist3 450")
mc_closestdist3_450 <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer450, 
                             method = "euclidean",
                             k = 3)
mc_closestdist3_450
closestdist3_450 <- mc_closestdist3_450$suggested_gcms
tictoc::toc()
```

```{r closestdist3_gcms500}
tictoc::tic("Closestdist3 500")
mc_closestdist3_500 <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = buffer500, 
                             method = "euclidean",
                             k = 3)
mc_closestdist3_500
closestdist3_500 <- mc_closestdist3_500$suggested_gcms
tictoc::toc()
```

## Results

The change in size highlights that every algorithm is very susceptible to area selection. K-means and Hierarchical Clustering were, respectivelly, the most stable algorithms. In terms of Species Distribution Modeling, we can say that the GCM selection is affected by the accessible area for the species (the M component of the BAM diagram).

```{r result_hclust}
rbind(sort(hclust_50 ), 
      sort(hclust_100),
      sort(hclust_150), 
      sort(hclust_200),
      sort(hclust_250), 
      sort(hclust_300),
      sort(hclust_350), 
      sort(hclust_400),
      sort(hclust_450), 
      sort(hclust_500))
```

```{r result_kmeans}
rbind(sort(kmeans_50 ), 
      sort(kmeans_100),
      sort(kmeans_150), 
      sort(kmeans_200),
      sort(kmeans_250), 
      sort(kmeans_300),
      sort(kmeans_350), 
      sort(kmeans_400),
      sort(kmeans_450), 
      sort(kmeans_500))
```

```{r result_closestdist}
list(sort(closestdist_50 ), 
     sort(closestdist_100),
     sort(closestdist_150), 
     sort(closestdist_200),
     sort(closestdist_250), 
     sort(closestdist_300),
     sort(closestdist_350), 
     sort(closestdist_400),
     sort(closestdist_450), 
     sort(closestdist_500))
```

```{r result_closestdist3}
rbind(sort(closestdist3_50 ), 
      sort(closestdist3_100),
      sort(closestdist3_150), 
      sort(closestdist3_200),
      sort(closestdist3_250), 
      sort(closestdist3_300),
      sort(closestdist3_350), 
      sort(closestdist3_400),
      sort(closestdist3_450), 
      sort(closestdist3_500))
```
