---
title: "Resolution Effect"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Resolution Effect}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::knitr}
editor_options: 
  markdown: 
    wrap: 100
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Is chooseGCM affected by resolution?

This Rmarkdown is part of the following article:

Esser, L.F., Bailly, D., Lima, M.R., Ré, R. 2025. chooseGCM: A Toolkit to Select General Circulation Models in R. Global Change Biology , 31(1), e70008. Available at: https://doi.org/10.1111/gcb.70008.

## Introduction

chooseGCM is a solution for GCMs selection in Climate Change research. We built this Rmarkdown as a way to test the properties of the methods underlying chooseGCM. Results from each function will be presented side by side with changing resolution, allowing better comparison. Here, we have already downloaded GCMs data.

This Rmarkdown will apply different algorithms implemented in chooseGCM in different resolutions. The aim is to understand if resolution changes will affect the GCMs subset selected by algorithms. We expect that results will not change with resolution, since the increase in resolution is just an interpolation of coarser resolution. As a 30 arcsec resolution will be used, we considered just a local scale due to computational and time limitations. We also restricted the variables used to bio1 and bio12 to restrict variation included with variables selection (see the proper Rmarkdown on the effects of variable selection for more information).

```{r libs, message=F, warning=F}
# Install and open CRAN libraries
for(x in c("devtools", "geodata", "terra", "tictoc") ){
  if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
  }
}

# Install and open chooseGCM
if (!require("chooseGCM", character.only = TRUE)) {
    devtools::install_github("luizesser/chooseGCM")
    library("chooseGCM", character.only = TRUE)
}

# Install and open caretSDM
if (!require("caretSDM", character.only = TRUE)) {
    devtools::install_github("luizesser/caretSDM")
    library("caretSDM", character.only = TRUE)
}
```

## Preparing data

### Import study_area

For this demonstration, we will use Araucaria angustifolia, the brazilian pine, as a study case. This flagship species is a key species for the Atlantic Rainforest biodiversity hotspot. Species records were obtained from GBIF and cleaned by excluding records from outside the native region and NAs. This set of records was then used to build a shape of accessible area by merging buffers of 500 km around each record.

```{r study_area}
occ_data <- geodata::sp_occurrence("Araucaria", "angustifolia")
occ_data <- occ_data |>
  dplyr::filter(lon >= -64) |> 
  dplyr::filter(lon <= -33) |> 
  dplyr::filter(lat <= -19.5) |> 
  dplyr::filter(lat >= -33)

study_area <- na.omit(occ_data[,c("lon", "lat")]) |>
  sf::st_as_sf(coords= c(1,2), crs = 4326) |> 
  sf::st_transform(crs=sf::st_crs(6933)) |>
  sf::st_buffer(dist = 500000) |> 
  sf::st_union()
```

### Download GCMs

To obtain GCMs data, we recommend using `geodata` package, which links directly to the WorldClim 2.1 database. We have previously downloaded GCMs data, but an example code is supplied below.

```{r download_gcms, eval=FALSE}
gcms <- geodata:::.cmods
for (r in c(10, 5, 2.5)) {
  for ( g in gcms ) {
    geodata::cmip6_world(model = g,
                         ssp = "585",
                         time = "2081-2100",
                         var = "bioc",
                         res = r,
                         path = ""
    )
  }
}
```

### Import GCMs

The function `import_gcms` helps the user to import data in the correct way, which is a list of GCMs. Each element of the list is a `SpatRaster` from `terra` package, containing all variables. The list should be named with GCMs names to better plot results.

```{r import_gcms10}
tictoc::tic()
s10 <- chooseGCM::import_gcms(path = "~/storage/WC_data/WC_data_all_gcms_10/", 
                              gcm_names = c("ACCESS-CM2","ACCESS-ESM1-5",
                                            "CanESM5-CanOE","CMCC-ESM2",
                                            "CNRM-CM6-1-HR","CNRM-CM6-1",
                                            "CNRM-ESM2-1","EC-Earth3-Veg",
                                            "EC-Earth3-Veg-LR","FIO-ESM-2-0",
                                            "GISS-E2-1-G","GISS-E2-1-H",
                                            "HadGEM3-GC31-LL","INM-CM5-0",
                                            "INM-CM4-8","IPSL-CM6A-LR",
                                            "MIROC-ES2L","MIROC6",
                                            "MPI-ESM1-2-LR","MPI-ESM1-2-HR",
                                            "MRI-ESM2-0","UKESM1-0-LL"))
tictoc::toc()
```

```{r import_gcms5}
tictoc::tic()
s5 <- chooseGCM::import_gcms(path = "~/storage/WC_data/WC_data_all_gcms_5", 
                              gcm_names = c("ACCESS-CM2","ACCESS-ESM1-5",
                                            "CanESM5-CanOE","CMCC-ESM2",
                                            "CNRM-CM6-1-HR","CNRM-CM6-1",
                                            "CNRM-ESM2-1","EC-Earth3-Veg",
                                            "EC-Earth3-Veg-LR","FIO-ESM-2-0",
                                            "GISS-E2-1-G","GISS-E2-1-H",
                                            "HadGEM3-GC31-LL","INM-CM5-0",
                                            "INM-CM4-8","IPSL-CM6A-LR",
                                            "MIROC-ES2L","MIROC6",
                                            "MPI-ESM1-2-LR","MPI-ESM1-2-HR",
                                            "MRI-ESM2-0","UKESM1-0-LL"))
tictoc::toc()
```

```{r import_gcms25}
tictoc::tic()
s25 <- chooseGCM::import_gcms(path = "~/storage/WC_data/WC_data_all_gcms_25", 
                              gcm_names = c("ACCESS-CM2","ACCESS-ESM1-5",
                                            "CanESM5-CanOE","CMCC-ESM2",
                                            "CNRM-CM6-1-HR","CNRM-CM6-1",
                                            "CNRM-ESM2-1","EC-Earth3-Veg",
                                            "EC-Earth3-Veg-LR","FIO-ESM-2-0",
                                            "GISS-E2-1-G","GISS-E2-1-H",
                                            "HadGEM3-GC31-LL","INM-CM5-0",
                                            "INM-CM4-8","IPSL-CM6A-LR",
                                            "MIROC-ES2L","MIROC6",
                                            "MPI-ESM1-2-LR","MPI-ESM1-2-HR",
                                            "MRI-ESM2-0","UKESM1-0-LL"))
tictoc::toc()
```

```{r import_gcms30}
tictoc::tic()
s30 <- chooseGCM::import_gcms(path = "~/storage/WC_data/WC_data_all_gcms_30", 
                              gcm_names = c("ACCESS-CM2","ACCESS-ESM1-5",
                                            "CanESM5-CanOE","CMCC-ESM2",
                                            "CNRM-CM6-1-HR","CNRM-CM6-1",
                                            "CNRM-ESM2-1","EC-Earth3-Veg",
                                            "EC-Earth3-Veg-LR","FIO-ESM-2-0",
                                            "GISS-E2-1-G","GISS-E2-1-H",
                                            "HadGEM3-GC31-LL","INM-CM5-0",
                                            "INM-CM4-8","IPSL-CM6A-LR",
                                            "MIROC-ES2L","MIROC6",
                                            "MPI-ESM1-2-LR","MPI-ESM1-2-HR",
                                            "MRI-ESM2-0","UKESM1-0-LL"))
tictoc::toc()
```

Importing GCMs data into R had an marginal difference in time, despite the big difference in size from different files.

## Applying algorithms

### Hierarchical Clustering selection

```{r hclust_gcms10}
tictoc::tic("Hierarchical Clustering subset s10")
mc_hclust10 <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = study_area, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "hclust")
mc_hclust10
hclust_subset10 <- mc_hclust10$suggested_gcms$k5
tictoc::toc()
```

```{r hclust_gcms5}
tictoc::tic("Hierarchical Clustering subset s5")
mc_hclust5 <- montecarlo_gcms(s5, 
                             var_names = c("bio1", "bio12"), 
                             study_area = study_area, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "hclust")
mc_hclust5
hclust_subset5 <- mc_hclust5$suggested_gcms$k5
tictoc::toc()
```

```{r hclust_gcms25}
tictoc::tic("Hierarchical Clustering subset s25")
mc_hclust25 <- montecarlo_gcms(s25, 
                             var_names = c("bio1", "bio12"), 
                             study_area = study_area, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "hclust")
mc_hclust25
hclust_subset25 <- mc_hclust25$suggested_gcms$k5
tictoc::toc()
```

```{r hclust_gcms30}
tictoc::tic("Hierarchical Clustering subset s30")
mc_hclust30 <- montecarlo_gcms(s30, 
                             var_names = c("bio1", "bio12"), 
                             study_area = study_area, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "hclust")
mc_hclust30
hclust_subset30 <- mc_hclust30$suggested_gcms$k5
tictoc::toc()
```

### Kmeans selection

```{r kmeans_gcms10}
tictoc::tic("K-means Clustering subset s10")
mc_kmeans10 <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = study_area, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "kmeans")
mc_kmeans10
kmeans_subset10 <- mc_kmeans10$suggested_gcms$k5
tictoc::toc()
```

```{r kmeans_gcms5}
tictoc::tic("K-means Clustering subset s5")
mc_kmeans5 <- montecarlo_gcms(s5, 
                             var_names = c("bio1", "bio12"), 
                             study_area = study_area, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "kmeans")
mc_kmeans5
kmeans_subset5 <- mc_kmeans5$suggested_gcms$k5
tictoc::toc()
```

```{r kmeans_gcms25}
tictoc::tic("K-means Clustering subset s25")
mc_kmeans25 <- montecarlo_gcms(s25, 
                             var_names = c("bio1", "bio12"), 
                             study_area = study_area, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "kmeans")
mc_kmeans25
kmeans_subset25 <- mc_kmeans25$suggested_gcms$k5
tictoc::toc()
```

```{r kmeans_gcms30}
tictoc::tic("K-means Clustering subset s30")
mc_kmeans30 <- montecarlo_gcms(s30, 
                             var_names = c("bio1", "bio12"), 
                             study_area = study_area, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "kmeans")
mc_kmeans30
kmeans_subset30 <- mc_kmeans30$suggested_gcms$k5
tictoc::toc()
```

### Closestdist selection

```{r closestdist_gcms10}
tictoc::tic("Closestdist subset s10")
mc_closestdist10 <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = study_area, 
                             method = "euclidean")
mc_closestdist10
closestdist_subset10 <- mc_closestdist10$suggested_gcms
tictoc::toc()
```

```{r closestdist_gcms5}
tictoc::tic("Closestdist subset s5")
mc_closestdist5 <- closestdist_gcms(s5, 
                             var_names = c("bio1", "bio12"), 
                             study_area = study_area, 
                             method = "euclidean")
mc_closestdist5
closestdist_subset5 <- mc_closestdist5$suggested_gcms
tictoc::toc()
```

```{r closestdist_gcms25}
tictoc::tic("Closestdist subset s25")
mc_closestdist25 <- closestdist_gcms(s25, 
                             var_names = c("bio1", "bio12"), 
                             study_area = study_area, 
                             method = "euclidean")
mc_closestdist25
closestdist_subset25 <- mc_closestdist25$suggested_gcms
tictoc::toc()
```

```{r closestdist_gcms30}
tictoc::tic("Closestdist subset s30")
mc_closestdist30 <- closestdist_gcms(s30, 
                             var_names = c("bio1", "bio12"), 
                             study_area = study_area, 
                             method = "euclidean")
mc_closestdist30
closestdist_subset30 <- mc_closestdist30$suggested_gcms
tictoc::toc()
```

### Closestdist selection with fixed k

```{r closestdist3_gcms10}
tictoc::tic("Closestdist3 subset s10")
mc_closestdist310 <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = study_area, 
                             method = "euclidean",
                             k = 3)
mc_closestdist310
closestdist3_subset10 <- mc_closestdist310$suggested_gcms
tictoc::toc()
```

```{r closestdist3_gcms5}
tictoc::tic("Closestdist3 subset s5")
mc_closestdist35 <- closestdist_gcms(s5, 
                             var_names = c("bio1", "bio12"), 
                             study_area = study_area, 
                             method = "euclidean",
                             k = 3)
mc_closestdist35
closestdist3_subset5 <- mc_closestdist35$suggested_gcms
tictoc::toc()
```

```{r closestdist3_gcms25}
tictoc::tic("Closestdist3 subset s25")
mc_closestdist325 <- closestdist_gcms(s25, 
                             var_names = c("bio1", "bio12"), 
                             study_area = study_area, 
                             method = "euclidean",
                             k = 3)
mc_closestdist325
closestdist3_subset25 <- mc_closestdist325$suggested_gcms
tictoc::toc()
```

```{r closestdist3_gcms30}
tictoc::tic("Closestdist3 subset s30")
mc_closestdist330 <- closestdist_gcms(s30, 
                             var_names = c("bio1", "bio12"), 
                             study_area = study_area, 
                             method = "euclidean",
                             k = 3)
mc_closestdist330
closestdist3_subset30 <- mc_closestdist330$suggested_gcms
tictoc::toc()
```

## Results

Contrary to our expectations, the change in resolution affected the outputs in every algorithm implemented. The only algorithm with a more stable result was the Kmeans Clustering.

```{r result_hclust}
table(c(hclust_subset10, hclust_subset5, hclust_subset25, hclust_subset30))
```

```{r result_kmeans}
table(c(kmeans_subset10, kmeans_subset5, kmeans_subset25, kmeans_subset30))
```

```{r result_closestdist}
table(c(closestdist_subset10, closestdist_subset5, closestdist_subset25, closestdist_subset30))
```

```{r result_closestdist3}
table(c(closestdist3_subset10, closestdist3_subset5, closestdist3_subset25, closestdist3_subset30))
```
