---
author: "Luíz Fernando Esser"
date: "2024-11-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# chooseGCM timespans

This Rmarkdown is part of the following article:

Esser, L.F., Bailly, D., Lima, M.R., Ré, R. 2025. chooseGCM: A Toolkit to Select General Circulation Models in R. Global Change Biology , 31(1), e70008. Available at: https://doi.org/10.1111/gcb.70008.

## Introduction

chooseGCM is a solution for GCMs selection in Climate Change research. We built this Rmarkdown as a way to test the properties of the methods underlying chooseGCM. Results from each function will be presented side by side with changing variables, allowing better comparison.

This RMarkdown was built upon tests made from reviewer 1 to inform regarding timespans to each function. The only function we will not be covering is the \code{worldclim_data}, which takes a lot of time to download data and the timespan depends majorly on the internet connection and not on the coding optimization.

```{r}
library(tidyverse)
library(sf)
library(chooseGCM)
library(tictoc)
```

# 10 min  ------------------------------------------------------------------

## import ----

```{r}
tictoc::tic()
s <- chooseGCM::import_gcms(path = "~/storage/WC_data/WC_data_all_gcms_10")
tictoc::toc()
s

names(s)
names(s) <- gsub("_ssp585_10_2090", "", names(s))
names(s)
```

## variable names and study area  ----

```{r}
var_names <- c("bio5", "bio13", "bio15")

study_area_parana <- geodata::gadm(country = "Brazil", path = "input_data/") %>% 
  sf::st_as_sf() %>% 
  dplyr::filter(NAME_1 == "Paraná")
study_area_parana

plot(study_area_parana$geometry)
```

## compare ----

```{r}
tictoc::tic()
res10 <- chooseGCM::compare_gcms(s, var_names, study_area_parana, k = 3) 
tictoc::toc()

res10$statistics_gcms
```

## summary ----

```{r}
tictoc::tic()
s_sum <- chooseGCM::summary_gcms(s, var_names, study_area_parana) 
tictoc::toc()

s_sum
```

## correlation ----

```{r}
tictoc::tic()
s_cor <- chooseGCM::cor_gcms(s, var_names, study_area_parana, method = "pearson") 
tictoc::toc()

s_cor
```

## distance ----

```{r}
tictoc::tic()
s_dist <- chooseGCM::dist_gcms(s, var_names, study_area_parana, method = "euclidean")
tictoc::toc()

s_dist
```

## k-means ----

```{r}
tictoc::tic()
chooseGCM::kmeans_gcms(s, var_names, study_area_parana, k = 3, method = "euclidean") 
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::kmeans_gcms(s, var_names, study_area_parana, k = 3)
tictoc::toc()
```

## hierarchical clustering ----

```{r}
tictoc::tic()
chooseGCM::hclust_gcms(s, var_names, study_area_parana, k = 3) 
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::hclust_gcms(s, var_names, study_area_parana, k = 3, n = 1000) 
tictoc::toc()
```

## Closestdist algorithm ----

```{r}
tictoc::tic()
chooseGCM::closestdist_gcms(s, var_names, study_area_parana, k = 3) 
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::closestdist_gcms(s, var_names, study_area_parana) 
tictoc::toc()
```

## number of clusters ----

```{r}
tictoc::tic()
chooseGCM::optk_gcms(s, var_names, study_area_parana, cluster = "kmeans", method = "wss", n = 1000) 
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::optk_gcms(s, var_names, study_area_parana, cluster = "kmeans", method = "silhouette", n = 1000) 
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::optk_gcms(s, var_names, study_area_parana, cluster = "kmeans", method = "gap_stat", n = 1000) 
tictoc::toc()
```

## monte carlo permutations ----

```{r}
tictoc::tic()
chooseGCM::montecarlo_gcms(s, var_names, study_area_parana, perm = 10000, method = "euclidean") 
tictoc::toc()
```

## environment ----

```{r}
tictoc::tic()
chooseGCM::env_gcms(s, var_names, study_area_parana, highlight = res10$suggested_gcms$k3) 
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::env_gcms(s, var_names, study_area_parana, highlight = "sum")
tictoc::toc()
```

# 5 min ------------------------------------------------------------------

## import ----

```{r}
tictoc::tic()
s <- chooseGCM::import_gcms(path = "~/storage/WC_data/WC_data_all_gcms_5") 
tictoc::toc()

s

names(s)
names(s) <- gsub("_ssp585_5_2090", "", names(s))
names(s)
```

## variable names and study area  ----

```{r}
var_names <- c("bio5", "bio13", "bio15")

study_area_parana <- geodata::gadm(country = "Brazil", path = "input_data/") %>% 
  sf::st_as_sf() %>% 
  dplyr::filter(NAME_1 == "Paraná")
study_area_parana

plot(study_area_parana$geometry)
```

## compare ----

```{r}
tictoc::tic()
res5 <- chooseGCM::compare_gcms(s, var_names, study_area_parana, k = 3) 
tictoc::toc()

res5$statistics_gcms
```

## summary ----

```{r}
tictoc::tic()
s_sum <- chooseGCM::summary_gcms(s, var_names, study_area_parana)
tictoc::toc()

s_sum
```

## correlation ----

```{r}
tictoc::tic()
s_cor <- chooseGCM::cor_gcms(s, var_names, study_area_parana, method = "pearson") 
tictoc::toc()

s_cor
```

## distance ----

```{r}
tictoc::tic()
s_dist <- chooseGCM::dist_gcms(s, var_names, study_area_parana, method = "euclidean") 
tictoc::toc()

s_dist
```

## k-means ----

```{r}
tictoc::tic()
chooseGCM::kmeans_gcms(s, var_names, study_area_parana, k = 3, method = "euclidean") 
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::kmeans_gcms(s, var_names, study_area_parana, k = 3) 
tictoc::toc()
```

## hierarchical clustering ----

```{r}
tictoc::tic()
chooseGCM::hclust_gcms(s, var_names, study_area_parana, k = 3) 
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::hclust_gcms(s, var_names, study_area_parana, k = 3, n = 1000) 
tictoc::toc()
```

## Closestdist algorithm ----

```{r}
tictoc::tic()
chooseGCM::closestdist_gcms(s, var_names, study_area_parana, k = 3) 
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::closestdist_gcms(s, var_names, study_area_parana) 
tictoc::toc()
```

## number of clusters ----

```{r}
tictoc::tic()
chooseGCM::optk_gcms(s, var_names, study_area_parana, cluster = "kmeans", method = "wss", n = 1000)
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::optk_gcms(s, var_names, study_area_parana, cluster = "kmeans", method = "silhouette", n = 1000)
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::optk_gcms(s, var_names, study_area_parana, cluster = "kmeans", method = "gap_stat", n = 1000)
tictoc::toc()
```

## monte carlo permutations ----

```{r}
tictoc::tic()
chooseGCM::montecarlo_gcms(s, var_names, study_area_parana, perm = 10000, method = "euclidean") 
tictoc::toc()
```

## environment ----

```{r}
tictoc::tic()
chooseGCM::env_gcms(s, var_names, study_area_parana, highlight = res5$suggested_gcms$k3) 
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::env_gcms(s, var_names, study_area_parana, highlight = "sum") 
tictoc::toc()
```

# 2.5 min ------------------------------------------------------------------

## import ----

```{r}
tictoc::tic()
s <- chooseGCM::import_gcms(path = "~/storage/WC_data/WC_data_all_gcms_25")
tictoc::toc()

s

names(s)
names(s) <- gsub("_ssp585_2.5_2090", "", names(s))
names(s)
```

## variable names and study area  ----

```{r}
var_names <- c("bio5", "bio13", "bio15")

study_area_parana <- geodata::gadm(country = "Brazil", path = "input_data/") %>% 
  sf::st_as_sf() %>% 
  dplyr::filter(NAME_1 == "Paraná")
study_area_parana

plot(study_area_parana$geometry)
```

## compare ----

```{r}
tictoc::tic()
res25 <- chooseGCM::compare_gcms(s, var_names, study_area_parana, k = 3) 
tictoc::toc()

res25$statistics_gcms
```

## summary ----

```{r}
tictoc::tic()
s_sum <- chooseGCM::summary_gcms(s, var_names, study_area_parana) 
tictoc::toc()

s_sum
```

## correlation ----

```{r}
tictoc::tic()
s_cor <- chooseGCM::cor_gcms(s, var_names, study_area_parana, method = "pearson")
tictoc::toc()

s_cor
```

## distance ----

```{r}
tictoc::tic()
s_dist <- chooseGCM::dist_gcms(s, var_names, study_area_parana, method = "euclidean")
tictoc::toc()

s_dist
```

## k-means ----

```{r}
tictoc::tic()
chooseGCM::kmeans_gcms(s, var_names, study_area_parana, k = 3, method = "euclidean")
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::kmeans_gcms(s, var_names, study_area_parana, k = 3) 
tictoc::toc()
```

## hierarchical clustering ----

```{r}
tictoc::tic()
chooseGCM::hclust_gcms(s, var_names, study_area_parana, k = 3) 
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::hclust_gcms(s, var_names, study_area_parana, k = 3, n = 1000) 
tictoc::toc()
```

## Closestdist algorithm ----

```{r}
tictoc::tic()
chooseGCM::closestdist_gcms(s, var_names, study_area_parana, k = 3) 
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::closestdist_gcms(s, var_names, study_area_parana) 
tictoc::toc()
```

## number of clusters ----

```{r}
tictoc::tic()
chooseGCM::optk_gcms(s, var_names, study_area_parana, cluster = "kmeans", method = "wss", n = 1000) 
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::optk_gcms(s, var_names, study_area_parana, cluster = "kmeans", method = "silhouette", n = 1000) 
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::optk_gcms(s, var_names, study_area_parana, cluster = "kmeans", method = "gap_stat", n = 1000) 
tictoc::toc()
```

## monte carlo permutations ----

```{r}
tictoc::tic()
chooseGCM::montecarlo_gcms(s, var_names, study_area_parana, perm = 10000, method = "euclidean") 
tictoc::toc()
```

## environment ----

```{r}
tictoc::tic()
chooseGCM::env_gcms(s, var_names, study_area_parana, highlight = res25$suggested_gcms$k3) 
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::env_gcms(s, var_names, study_area_parana, highlight = "sum")
tictoc::toc()
```

# 30 sec ------------------------------------------------------------------

## import ----

```{r}
tictoc::tic()
s <- chooseGCM::import_gcms(path = "~/storage/WC_data/WC_data_all_gcms_30") 
tictoc::toc()

s

names(s)
names(s) <- gsub("_ssp585_30_2090", "", names(s))
names(s)
```

## variable names and study area  ----

```{r}
var_names <- c("bio5", "bio13", "bio15")

study_area_parana <- geodata::gadm(country = "Brazil", path = "input_data/") %>% 
  sf::st_as_sf() %>% 
  dplyr::filter(NAME_1 == "Paraná")
study_area_parana

plot(study_area_parana$geometry)
```

## compare ----

```{r}
tictoc::tic()
res30 <- chooseGCM::compare_gcms(s, var_names, study_area_parana, k = 3) 
tictoc::toc()

res30$statistics_gcms
```

## summary ----

```{r}
tictoc::tic()
s_sum <- chooseGCM::summary_gcms(s, var_names, study_area_parana) 
tictoc::toc()

s_sum
```

## correlation ----

```{r}
tictoc::tic()
s_cor <- chooseGCM::cor_gcms(s, var_names, study_area_parana, method = "pearson") 
tictoc::toc()

s_cor
```

## distance ----

```{r}
tictoc::tic()
s_dist <- chooseGCM::dist_gcms(s, var_names, study_area_parana, method = "euclidean") 
tictoc::toc()

s_dist
```

## k-means ----

```{r}
tictoc::tic()
chooseGCM::kmeans_gcms(s, var_names, study_area_parana, k = 3, method = "euclidean") 
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::kmeans_gcms(s, var_names, study_area_parana, k = 3) 
tictoc::toc()
```

## hierarchical clustering ----

```{r}
tictoc::tic()
chooseGCM::hclust_gcms(s, var_names, study_area_parana, k = 3) 
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::hclust_gcms(s, var_names, study_area_parana, k = 3, n = 1000) 
tictoc::toc()
```

## Closestdist algorithm ----

```{r}
tictoc::tic()
chooseGCM::closestdist_gcms(s, var_names, study_area_parana, k = 3) 
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::closestdist_gcms(s, var_names, study_area_parana) 
tictoc::toc()
```

## number of clusters ----

```{r}
tictoc::tic()
chooseGCM::optk_gcms(s, var_names, study_area_parana, cluster = "kmeans", method = "wss", n = 1000) 
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::optk_gcms(s, var_names, study_area_parana, cluster = "kmeans", method = "silhouette", n = 1000) 
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::optk_gcms(s, var_names, study_area_parana, cluster = "kmeans", method = "gap_stat", n = 1000) 
tictoc::toc()
```

## monte carlo permutations ----

```{r}
tictoc::tic()
chooseGCM::montecarlo_gcms(s, var_names, study_area_parana, perm = 10000, method = "euclidean") 
tictoc::toc()
```

## environment ----

```{r}

tictoc::tic()
chooseGCM::env_gcms(s, var_names, study_area_parana, highlight = res30$suggested_gcms$k3) 
tictoc::toc()
```

```{r}
tictoc::tic()
chooseGCM::env_gcms(s, var_names, study_area_parana, highlight = "sum") 
tictoc::toc()
```

# end ---------------------------------------------------------------------
