---
author: "Luíz Fernando Esser"
date: "2024-11-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Is chooseGCM affected by variable selection?

This Rmarkdown is part of the following article:

Esser, L.F., Bailly, D., Lima, M.R., Ré, R. 2025. chooseGCM: A Toolkit to Select General Circulation Models in R. Global Change Biology , 31(1), e70008. Available at: https://doi.org/10.1111/gcb.70008.

## Introduction

chooseGCM is a solution for GCMs selection in Climate Change research. We built this Rmarkdown as a way to test the properties of the methods underlying chooseGCM. Results from each function will be presented side by side with changing variables, allowing better comparison. Here, we have already downloaded GCMs data.

This Rmarkdown will apply different algorithms implemented in chooseGCM in different variables sets. The aim is to understand if different variables sets will affect the GCMs subset selected by algorithms. We expect that results will change with varibles sets, since this change will change the very core of the comparison process, which is values presented. To showcase this, we used four sets of bioclimatic variables: (1) Bio1 and Bio12, which are the main temperature and precipitation variables (Annual Mean Temperature and Annual Precipitation, respectively); (2) Temperature variables (Bio1 to Bio11); (3) Precipitation variables (Bio12 to Bio19); and (4) all bioclimatic variables.

```{r libs, message=F, warning=F}
# Install and open CRAN libraries
for(x in c("devtools", "geodata", "terra", "tictoc") ){
  if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
  }
}

# Install and open chooseGCM
if (!require("chooseGCM", character.only = TRUE)) {
    devtools::install_github("luizesser/chooseGCM")
    library("chooseGCM", character.only = TRUE)
}

# Install and open caretSDM
if (!require("caretSDM", character.only = TRUE)) {
    devtools::install_github("luizesser/caretSDM")
    library("caretSDM", character.only = TRUE)
}
```

## Preparing data

### Import study_area

For this demonstration, we will use Araucaria angustifolia, the brazilian pine, as a study case. This flagship species is a key species for the Atlantic Rainforest biodiversity hotspot. Species records were obtained from GBIF and cleaned by excluding records from outside the native region and NAs. This set of records was then used to build a shape of accessible area by merging buffers of 500 km around each record.

```{r study_area}
occ_data <- geodata::sp_occurrence("Araucaria", "angustifolia")
occ_data <- occ_data |>
  dplyr::filter(lon >= -64) |> 
  dplyr::filter(lon <= -33) |> 
  dplyr::filter(lat <= -19.5) |> 
  dplyr::filter(lat >= -33)

study_area <- na.omit(occ_data[,c("lon", "lat")]) |>
  sf::st_as_sf(coords= c(1,2), crs = 4326) |> 
  sf::st_transform(crs=sf::st_crs(6933)) |>
  sf::st_buffer(dist = 500000) |> 
  sf::st_union()
```

### Download GCMs

To obtain GCMs data, we recommend using `geodata` package, which links directly to the WorldClim 2.1 database. We have previously downloaded GCMs data, but an example code is supplied below.

```{r download_gcms, eval=FALSE}
gcms <- geodata:::.cmods
for (r in c(10)) {
  for ( g in gcms ) {
    geodata::cmip6_world(model = g,
                         ssp = "585",
                         time = "2081-2100",
                         var = "bioc",
                         res = r,
                         path = ""
    )
  }
}
```

### Import GCMs

The function `import_gcms` helps the user to import data in the correct way, which is a list of GCMs. Each element of the list is a `SpatRaster` from `terra` package, containing all variables. The list should be named with GCMs names to better plot results.

```{r import_gcms10}
tictoc::tic()
s10 <- chooseGCM::import_gcms(path = "~/storage/WC_data/WC_data_all_gcms_10/", 
                              gcm_names = c("ACCESS-CM2","ACCESS-ESM1-5",
                                            "CanESM5-CanOE","CMCC-ESM2",
                                            "CNRM-CM6-1-HR","CNRM-CM6-1",
                                            "CNRM-ESM2-1","EC-Earth3-Veg",
                                            "EC-Earth3-Veg-LR","FIO-ESM-2-0",
                                            "GISS-E2-1-G","GISS-E2-1-H",
                                            "HadGEM3-GC31-LL","INM-CM5-0",
                                            "INM-CM4-8","IPSL-CM6A-LR",
                                            "MIROC-ES2L","MIROC6",
                                            "MPI-ESM1-2-LR","MPI-ESM1-2-HR",
                                            "MRI-ESM2-0","UKESM1-0-LL"))
tictoc::toc()
```

## Applying algorithms

First we will set a seed to standardize randomic outputs.

```{r seed}
set.seed(1)
```

Now we will perform the Monte Carlo analysis to each type of algorithm and to different sets of variables.

### Hierarchical Clustering selection

```{r hclust_gcms10}
tictoc::tic("Hierarchical Clustering Main Bioclimatics")
mc_hclust_main <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = study_area, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "hclust")
mc_hclust_main
hclust_main <- mc_hclust_main$suggested_gcms$k5
tictoc::toc()
```

```{r hclust_gcms5}
tictoc::tic("Hierarchical Clustering Temperature Variables")
mc_hclust_temp <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio2", "bio3", "bio4", "bio5", "bio6", "bio7", "bio8", "bio9", "bio10", "bio11"), 
                             study_area = study_area, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "hclust")
mc_hclust_temp
hclust_temp <- mc_hclust_temp$suggested_gcms$k5
tictoc::toc()
```

```{r hclust_gcms25}
tictoc::tic("Hierarchical Clustering Precipitation Variables")
mc_hclust_prec <- montecarlo_gcms(s10, 
                             var_names = c("bio12", "bio13", "bio14", "bio15", "bio16", "bio17", "bio18", "bio19"), 
                             study_area = study_area, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "hclust")
mc_hclust_prec
hclust_prec <- mc_hclust_prec$suggested_gcms$k5
tictoc::toc()
```

```{r hclust_gcms30}
tictoc::tic("Hierarchical Clustering All Variables")
mc_hclust_all <- montecarlo_gcms(s10, 
                             var_names = "all", 
                             study_area = study_area, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "hclust")
mc_hclust_all
hclust_all <- mc_hclust_all$suggested_gcms$k5
tictoc::toc()
```

### Kmeans selection

```{r kmeans_gcms10}
tictoc::tic("K-means Clustering Main Variables")
mc_kmeans_main <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = study_area, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "kmeans")
mc_kmeans_main
kmeans_main <- mc_kmeans_main$suggested_gcms$k5
tictoc::toc()
```

```{r kmeans_gcms5}
tictoc::tic("K-means Clustering Temperature Variables")
mc_kmeans_temp <- montecarlo_gcms(s10, 
                             var_names = c("bio1", "bio2", "bio3", "bio4", "bio5", "bio6", "bio7", "bio8", "bio9", "bio10", "bio11"), 
                             study_area = study_area, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "kmeans")
mc_kmeans_temp
kmeans_temp <- mc_kmeans_temp$suggested_gcms$k5
tictoc::toc()
```

```{r kmeans_gcms25}
tictoc::tic("K-means Clustering Precipitation Variables")
mc_kmeans_prec <- montecarlo_gcms(s10, 
                             var_names = c("bio12", "bio13", "bio14", "bio15", "bio16", "bio17", "bio18", "bio19"), 
                             study_area = study_area, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "kmeans")
mc_kmeans_prec
kmeans_prec <- mc_kmeans_prec$suggested_gcms$k5
tictoc::toc()
```

```{r kmeans_gcms30}
tictoc::tic("K-means Clustering All Variables")
mc_kmeans_all <- montecarlo_gcms(s10, 
                             var_names = "all", 
                             study_area = study_area, 
                             scale = TRUE,
                             perm = 10000, 
                             dist_method = "euclidean", 
                             clustering_method = "kmeans")
mc_kmeans_all
kmeans_all <- mc_kmeans_all$suggested_gcms$k5
tictoc::toc()
```

### Closestdist selection

```{r closestdist_gcms10}
tictoc::tic("Closestdist Main Variables")
mc_closestdist_main <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = study_area, 
                             method = "euclidean")
mc_closestdist_main
closestdist_main <- mc_closestdist_main$suggested_gcms
tictoc::toc()
```

```{r closestdist_gcms5}
tictoc::tic("Closestdist Temperature Variables")
mc_closestdist_temp <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio2", "bio3", "bio4", "bio5", "bio6", "bio7", "bio8", "bio9", "bio10", "bio11"), 
                             study_area = study_area, 
                             method = "euclidean")
mc_closestdist_temp
closestdist_temp <- mc_closestdist_temp$suggested_gcms
tictoc::toc()
```

```{r closestdist_gcms25}
tictoc::tic("Closestdist Precipitation Variables")
mc_closestdist_prec <- closestdist_gcms(s10, 
                             var_names = c("bio12", "bio13", "bio14", "bio15", "bio16", "bio17", "bio18", "bio19"), 
                             study_area = study_area, 
                             method = "euclidean")
mc_closestdist_prec
closestdist_prec <- mc_closestdist_prec$suggested_gcms
tictoc::toc()
```

```{r closestdist_gcms30}
tictoc::tic("Closestdist All Variables")
mc_closestdist_all <- closestdist_gcms(s10, 
                             var_names = "all", 
                             study_area = study_area, 
                             method = "euclidean")
mc_closestdist_all
closestdist_all <- mc_closestdist_all$suggested_gcms
tictoc::toc()
```

### Closestdist selection with fixed k

```{r closestdist3_gcms10}
tictoc::tic("Closestdist3 Main Variables")
mc_closestdist3_main <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio12"), 
                             study_area = study_area, 
                             method = "euclidean",
                             k = 3)
mc_closestdist3_main
closestdist3_main <- mc_closestdist3_main$suggested_gcms
tictoc::toc()
```

```{r closestdist3_gcms5}
tictoc::tic("Closestdist3 Temperature Variables")
mc_closestdist3_temp <- closestdist_gcms(s10, 
                             var_names = c("bio1", "bio2", "bio3", "bio4", "bio5", "bio6", "bio7", "bio8", "bio9", "bio10", "bio11"), 
                             study_area = study_area, 
                             method = "euclidean",
                             k = 3)
mc_closestdist3_temp
closestdist3_temp <- mc_closestdist3_temp$suggested_gcms
tictoc::toc()
```

```{r closestdist3_gcms25}
tictoc::tic("Closestdist3 Precipitation Variables")
mc_closestdist3_prec <- closestdist_gcms(s10, 
                             var_names = c("bio12", "bio13", "bio14", "bio15", "bio16", "bio17", "bio18", "bio19"), 
                             study_area = study_area, 
                             method = "euclidean",
                             k = 3)
mc_closestdist3_prec
closestdist3_prec <- mc_closestdist3_prec$suggested_gcms
tictoc::toc()
```

```{r closestdist3_gcms30}
tictoc::tic("Closestdist3 All Variables")
mc_closestdist3_all <- closestdist_gcms(s10, 
                             var_names = "all", 
                             study_area = study_area, 
                             method = "euclidean",
                             k = 3)
mc_closestdist3_all
closestdist3_all <- mc_closestdist3_all$suggested_gcms
tictoc::toc()
```

## Results

As we expected, the change in variables affected the outputs in every algorithm implemented. It is possible to see that in the following chunks it was rare a GCM to be selected more than 2 times in the same algorithm.

```{r result_hclust}
table(c(hclust_main, hclust_temp, hclust_prec, hclust_all))
```

```{r result_kmeans}
table(c(kmeans_main, kmeans_temp, kmeans_prec, kmeans_all))
```

```{r result_closestdist}
table(c(closestdist_main, closestdist_temp, closestdist_prec, closestdist_all))
```

```{r result_closestdist3}
table(c(closestdist3_main, closestdist3_temp, closestdist3_prec, closestdist3_all))
```
